---
title: "Exploring the Patch Model"
author: "Russell Dinnage"
date: "15/08/2014"
output: 
  html_document:
    highlight: pygments
bibliography: "/home/din02g/Google Drive/References/PGHG.bib"
---

```{r knitr_init, echo=FALSE, results="asis", cache=FALSE}
library(knitr)

## Global options
options(max.print="75")
opts_chunk$set(fig.path="out/",
               echo=TRUE,
               cache=TRUE,
               cache.path="cache2/",
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               autodep = TRUE)
opts_knit$set(width=75)
```

Here, I document some testing of a stochastic patch model, which will be used in conjunction with a phylogenetic simulation model, to look at the effects of phylogenetic diversity in a resource species, on the diversity of consumer species. The patch model is very simple. It consists of a number of sites that exist in a patch. Each site is occupied by a fixed resource species. In this model, resource species cannot go extinct --- they are properties of the patch. This makes sense because we are mainly interested in the short-term local effect of diversity (rather than the long-term co-evolutionary / metacommunity effects).

```{r setup}
library(PGHGproject)
```